Exploring the Role of Feedback in Machine Learning Jupyter Notebooks

### 2024-07-01 report/emse24.tex
feat(report): emse24 updates

  \todo{update the numbers here}
Developing and testing ML systems is a highly interactive process which demands a human-in-the-loop approach. In addition to writing tests for the code base, practitioners are required to analyse and interpret several visualisations using their domain expertise to validate if an ML system satisfies the required set of functional and non-functional properties. Visualisations are frequently used to qualitatively assess various parts of an ML pipeline. However, implicit knowledge gained from visualisations must be translated to explicit analytical tests that fail when there is change in any component of the ML pipeline. We conduct an empirical analysis of Jupyter notebooks to catalogue the state-of-the-art mappings between ML visualisations and assertions. We mine Github to collect 54K Jupyter Notebooks that contain assertions written in Python. We develop a novel methodology to identify 1764 notebooks which contain an assertion below a visualisation. We manually analyse the 1.7K notebooks and identify 269 visualisation-assertion pairs that are semantically related to one another. We further investigate the 269 visualisation-assertion pairs and identify three frequently occurring testing patterns. We perform an in-depth analysis of 34 visualisation-assertion pairs and find that the assertions often fail to capture all the information present in their visual counter-part. Empirical evidence obtained in this study indicates that current software testing methods fail to address the unique challenges of ML. And emphasises the need for automated tools that bridge the gap between visual assessments and analytical assertions.

\keywords{ML Testing \and SE4AI \and Visualisations \and Assertions \and Computational Notebooks}
\end{abstract}

@@

@@

@@

% TODO: some sort of introduction to the following sub-sections here; we need to explain that for code cell outputs we try to show the python statement that produces the output along with the actual output. Somethimes actual output is not possible (eg. P68) because the statement may be inside an if statement.

@@

25K (8.5\%) notebooks contained at least one assertion. From the 8.5\% notebooks, we collected 89.6K assertions. 4.8K assertions (5\%) used methods from external testing libraries. Majority of the assertions however (85K or 95\%) were written using the \lstinline{assert} statement provided by the built-in Python standard library. We removed duplicate assertions for the remainder of the analysis, thus resulting in 27.1K assertions.

@@

\begin{table}
  \centering
  \caption{Examples of the most common tests performed in assert statements.}
  \begin{tabular}{@{}m{0.3\textwidth} m{0.3\textwidth}@{}}
    \toprule
    \emph{\textbf{Comparison Check}}&
    \emph{\textbf{Function Call}}\\
    \midrule

@@

Of the 95\% assert statements, 28.4K (31.7\%) have a failure message. Figure~\ref{fig:common-assert-msgs} shows the most frequent bi-grams that appear in these failure messages. Similar to the analysis presented above, we removed the duplicate messages prior to analysing the bi-grams resulting in 5.9K unique messages. The bi-grams indicate that the shape and data type of data structures are frequently verified.

@@

We collected 1.4M print statements from 180K notebooks (60\% of the total notebooks analysed in this study). Unlike assert statements, print statements do not have a fixed lexical semantic since the Python ast module interprets them as regular function calls. Additionally, print statements can be written in several forms---the \lstinline{print} function may have multiple arguments, any form of string interpolation supported by Python and output from function calls. Figure~\ref{fig:common-print-nodes} shows that the most common nodes appear inside of print statements are constants followed by function calls. Figure~\ref{fig:common-print-constant-types} further validates that the constants are almost entirely of type string. Based on these observations, we disect the print statements into two components---text written in natural language (from the \emph{Constant} nodes) and code (from all other nodes). Figure~\ref{fig:common-print-constants} and Figure~\ref{fig:common-print-not-constants} present the most common bi-grams that appear in the natural text and code of print statements respectively. Both figures collectively indicate that print statements are frequently used to print various performance metrics of ML models.

@@

@@

We collected 1M last statements from 119K notebooks. Figure~\ref{fig:common-last-nodes} shows that the most frequent type of code written as the final statement in a code cell is a function call. We further analyse the functions calls. Figure~\ref{fig:common-last-modules}

% most common functions?

@@

@@

@@

>>>>

  % TODO: update the abstract
  Developing and testing ML systems is a highly interactive process which demands a human-in-the-loop approach. In addition to writing tests for the code base, practitioners are required to analyse and interpret several visualisations using their domain expertise to validate if an ML system satisfies the required set of functional and non-functional properties. Visualisations are frequently used to qualitatively assess various parts of an ML pipeline. However, implicit knowledge gained from visualisations must be translated to explicit analytical tests that fail when there is change in any component of the ML pipeline. We conduct an empirical analysis of Jupyter notebooks to catalogue the state-of-the-art mappings between ML visualisations and assertions. We mine Github to collect 54K Jupyter Notebooks that contain assertions written in Python. We develop a novel methodology to identify 1764 notebooks which contain an assertion below a visualisation. We manually analyse the 1.7K notebooks and identify 269 visualisation-assertion pairs that are semantically related to one another. We further investigate the 269 visualisation-assertion pairs and identify three frequently occurring testing patterns. We perform an in-depth analysis of 34 visualisation-assertion pairs and find that the assertions often fail to capture all the information present in their visual counter-part. Empirical evidence obtained in this study indicates that current software testing methods fail to address the unique challenges of ML. And emphasises the need for automated tools that bridge the gap between visual assessments and analytical assertions.

\end{abstract}

\keywords{ML Testing \and SE4AI \and Visualisations \and Assertions \and Computational Notebooks}

@@

% TODO: feedback from Diomidis
% - In the introduction clarify (and perhaps illustrate) the life cycle and the role of notebooks.  Are these used to develop code that is then integrated into scripts, or is the notebook the execution vehicle, or is it informal throw-away code and rough notes on how to repeat a process in the future?  This will also inform the discussion and the implications.  Consider describing
%  a) the current development model, with the role of assertions, implicit checks, and visualization,
%  b) its limitations (which provide motivation for the study),
%  c) a sketch for an improved development model, and
%  d) the knowledge gaps associated with the improved model and how this study fills them.
%
% - The part starting "However, these decisions are often based on the conditions present when the model was initially created", which provides the motivation must be argued more convincingly.  I think that the motivation is not so much what is described in that paragraph, as the fact that, in the absence of detailed, up-to-date, and accurate requirements, rationale, and design documentation as well as tests, many of these can be derived from notebook interactions.

@@

% TODO: feedback from Diomidis
% - In Section 2, rather than jumping to literate programming, it would be more appropriate to discuss the CLI interaction style, which also encompasses Jupyter.
% - Section 2.2 (Python AST) can be shortened, shortened and moved to methods, or removed. Consider also describing the format of Jupyter notebooks.

@@

% TODO: feedback from Diomidis
% - Sections 4.1 and 4.2 are very interesting but overly detailed. I suggest to move most of the content to an appendix or an online appendix, and only present the uses in a table or a taxonomy tree.  In addition, consider moving the more general observations into a discussion section.
% - Consider starting with RQ2 (making it RQ1), because assertions are a concrete feature of Python code.  Then move to RQ1, which deals with implicit assertions.
% - I am afraid I cannot understand RQ3; consider revising it.  I would think that visualizations are verified visually, rather than through assertions.  You may have encountered some parts, but I would assume that the assertions mostly deal with things that can be more accurately and effectively be expressed as a predicate rather than viewed in a chart.
% TODO: some sort of introduction to the following sub-sections here; we need to explain that for code cell outputs we try to show the python statement that produces the output along with the actual output. Somethimes actual output is not possible (eg. P68) because the statement may be inside an if statement.

@@

25 thousand (8.5\%) notebooks contained at least one assertion. From the 8.5\% notebooks, we collected 89.6 thousand assertions. 4.8 thousand assertions (5\%) used methods from external testing libraries. 85 thousand assertions (95\%) however were written using the \lstinline{assert} statement provided by the built-in Python standard library. We removed duplicate assertions for the remainder of the analysis, thus resulting in 27.1 thousand assertions.

@@

\begin{table}
  \centering
  \caption{Examples of the most common tests performed in assert statements.}
  \begin{tabular}{@{}m{0.5\textwidth} m{0.5\textwidth}@{}}
    \toprule
    \emph{\textbf{Comparison Check}}&
    \emph{\textbf{Function Call}}\\
    \midrule

@@

Of the 95\% assert statements, 28.4 thousand (31.7\%) have a failure message. Figure~\ref{fig:common-assert-msgs} shows the most frequent bi-grams that appear in these failure messages. Similar to the analysis presented above, we removed the duplicate messages prior to analysing the bi-grams resulting in 5.9 thousand unique messages. The bi-grams indicate that the shape and data type of data structures are frequently verified.

@@

We collected 1.4 million print statements from 180 thousand notebooks (60\% of the total notebooks analysed in this study). Unlike assert statements, print statements do not have a fixed lexical semantic since the Python ast module interprets them as regular function calls. Additionally, print statements can be written in several forms---the \lstinline{print} function may have multiple arguments, any form of string interpolation supported by Python or output from function calls. Figure~\ref{fig:common-print-nodes} shows that the most common nodes inside of print statements are constants followed by function calls. Figure~\ref{fig:common-print-constant-types} further validates that the constants are almost entirely of type string. Based on these observations, we disect the print statements into two components---text written in natural language (from the \emph{Constant} nodes) and code (from all other nodes). Figure~\ref{fig:common-print-constants} and Figure~\ref{fig:common-print-not-constants} present the most common bi-grams that appear in the natural text and code of print statements respectively. Both figures collectively indicate that print statements are frequently used to print various performance metrics of ML models.

@@

We collected 1 million last statements from 119 thousand notebooks (40\% of the total notebooks analysed in this study). Figure~\ref{fig:common-last-nodes} shows that the most frequent type of code written as the final statement in a code cell is a function call. We further analyse the functions calls in more detail. Figure~\ref{fig:common-last-modules} and Figure~\ref{fig:common-last-functions} show the most modules and functions used inside the function calls respectively. The figures highlight the prominence of data visualisation libraries seaborn (\lstinline{sns}) and matplotlib (\lstinline{plt}) and visualisation functions such as \lstinline{plot} and \lstinline{countplot} in the notebooks. The data analysis tool pandas (represented by \lstinline{df} and \lstinline{pd}) also features prominently with many data exploration and preprocessing functions such as \lstinline{head}, \lstinline{value_counts} and \lstinline{describe} being used frequently. The presence of \lstinline{train} and \lstinline{model} modules along with the \lstinline{fit} function indicates the use of ML libraries.

@@

% TODO: we need better terminology here
\begin{figure}
  \centering
  \begin{minipage}{0.49\textwidth}
    \centering
    \includegraphics[width=\linewidth]{common-last-modules.pdf}
    \caption{Most common modules used in the function calls of last statements.}
    \label{fig:common-last-modules}
  \end{minipage}
  \hfill
  \begin{minipage}{0.49\textwidth}
    \centering
    \includegraphics[width=\linewidth]{common-last-functions.pdf}
    \caption{Most common functions used in the function calls of last statements.}
    \label{fig:common-last-functions}
  \end{minipage}
\end{figure}

\begin{figure}
  \centering
  \begin{minipage}{0.49\textwidth}
    \centering
    \includegraphics[width=\linewidth]{common-last-df-functions.pdf}
    \caption{Most common functions called on pandas data frames.}
    \label{fig:common-last-df-functions}
  \end{minipage}
  \hfill
  \begin{minipage}{0.49\textwidth}
    \centering
    \includegraphics[width=\linewidth]{common-last-visualisation-functions.pdf}
    \caption{Most common visualisation functions used in last statements.}
    \label{fig:common-last-visualisation-functions}
  \end{minipage}
\end{figure}

We further analysed the most common functions called on pandas data frames by isolating the last statements where the module is \lstinline{data} or \lstinline{df}. This is shown in Figure~\ref{fig:common-last-df-functions} which reveals a focus on exploratory data analysis and statistical operations. The frequent use of functions such as \lstinline{value_counts}, \lstinline{head}, and \lstinline{unique}, indicate an emphasis on understanding the structure and content of the data. Visualisation functions like \lstinline{plot} and \lstinline{hist} suggest that practitioners often create quick visualisations directly from data frames. Descriptive statistics are also prominent, with functions like \lstinline{describe}, \lstinline{mean}, and \lstinline{sum} being commonly used.

Similarly, we also analysed the most common visualisation functions used in notebooks by isolating the last statements where the module is \lstinline{sns} or \lstinline{plt}. This is shown in Figure~\ref{fig:common-last-visualisation-functions} which indicates that categorical data visualisation appears to be a primary focus, with \lstinline{countplot} and \lstinline{barplot} ranking high among the most frequently used functions. The prominence of \lstinline{heatmap} and \lstinline{boxplot} functions indicates a regular need for visualising correlations and statistical distributions. For continuous data, functions like \lstinline{distplot}, \lstinline{scatterplot}, and \lstinline{scatter} are commonly used. The presence of \lstinline{imshow} suggests that image-related visualisations are also significant in many notebooks.

@@

% TODO: feedback from Diomidis
% - "it represents an emerging testing practice within the ML community": I have commented before that assertions are different from tests. Assertions are a mechanism to increase the code's robustness, similar to type checks, array bounds checks, and thrown API exceptions.

@@

% TODO: feedback from Diomidis
% - The implications section focuses implicitly on assertions, at the expense of the other elements discussed earlier.  Also, it puts excessive emphasis on testing, which IIUC is not the paper's topic.

@@

% TODO: feedback from Diomidis
% - In Section 8 consider adding material regarind interactive ML development.
